---
title: "midterm_1_cheat_sheet"
output: html_document
date: "2024-02-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Lab 1.1

We create vectors using the `c` command. The `c` stands for concatenate. to make a vector:
```{r}
#_ <- c()
```

to download the 'tidyverse' package
```{r}
#library("tidyverse")
```


## Lab 2.1 & 2.2

### Types of Data
There are five frequently used `classes` of data: 1. numeric, 2. integer, 3. character, 4. logical, 5. complex.
```{r}
#my_numeric <- 42
#my_integer <- 2L #adding an L automatically denotes an integer
#my_character <- "universe"
#my_logical <- FALSE
#my_complex <- 2+4i
```

A numeric vector
```{r}
#my_vector <- c(10, 20, 30)
```

A character vector. Characters always have quotes and may be referred to as "strings".
```{r}
#days_of_the_week <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")
```

A convenient trick for creating a vector is to generate a sequence of numbers.
```{r}
#my_vector_sequence <- c(1:100)
```


To find out what type of data you are working with, use the `class()` function
```{r}
#class()
```

You can use the `is()` and `as()` functions to clarify or specify a type of data.
```{r}
#is.integer() #is something an integer?
```

```{r}
#as.integer() #create a new object specified as an integer
```

### Missing Data
`is.na` or `anyNA` are useful functions when dealing with NAs in data
```{r}
#is.na()
#anyNA()
#?na.rm=T #na.rm=T removes the NA values in the vector
```

### Data Matrices
Create a matrix using the `matrix()` command. We need to tell R how to organize the your vector using the `nrow` and `byrow` commands.
```{r}
#matrix <- matrix(vector, nrow = a_number, byrow = T)
```

Name the columns using `colnames()` using your column vector.
```{r}
#colnames(matrix) <- vector
```

Name the rows using `rownames()` using your row vector.
```{r}
#rownames(matrix) <- vector
```

can calculate the sum of one of the rows using 'rowSums'
```{r}
#rowSums()
```

can calculate the sum of one of the columns using 'colSums'
```{r}
#colSums()
```

To add a new row to a data matrix or data frame, use 'rbind'
```{r}
#rbind(matrix_or_frame, new_row_object_name)
```

to add a new column to a data matrix or data frame, use 'cbind'
```{r}
#cbind(matrix_or_frame, new_column_object_name)
```

The same methods of selecting elements in a vector apply to data matrices. We use `[]`. Ex: select the value in the first column and second row:
```{r}
#matrix[2,1] #Adding a colon `:` selects the specified elements in a column.
# can select elements in a whole row by putting nothing in the column area: [ ,row] and inversely [column, ]
```

To take the mean of all the rows, use 'rowMeans'
```{r}
#rowMeans()
```


## Lab 3.1

### Data Frames 
includes example for make a "data.frame()' from scratch

The `dim()` and `str()` commands provide the dimensions and/or structure of a data frame.
```{r}
#dim()
#str()
```

to change the names of data frame variables
```{r}
#data.frame(new_name=old_name) #or can use 'janitor' package to automatically correct all names
```

can also use 'rename()'
```{r}
#rename()
```


in addition to [], can also select values in an entire column using the `$` sign. it can also be used for adding more data to the data frame
```{r}
#data_frame$col_name
```

## Writing Data to File
There are many ways to save data in R, here we write our data frame to a csv file. We use `row.names = FALSE` to avoid row numbers from printing out. 
```{r}
#write.csv(data_frame, "data_frame.csv", row.names = FALSE)
```

## Lab 3.2

To import any file, first make sure that you are in the correct working directory. If you are not in the correct directory, R will not "see" the file.
```{r}
#getwd()
```

### Load the data
Here we open a .csv file. Since we are using the tidyverse, we open the file using `read_csv()`.
```{r}
#read.csv("file_name.csv") # '../' tells it to go farther back into my repository folder, if the data file is not located in the Lab folder 
```

```{r}
#readr::read_csv()
```


Can change a class from a character to a factor. A factor is when you have repeated sets of a type; categories of characters
```{r}
#data_frame$col_name <- as.factor(data_frame$col_name)
```

### Summary functions
We can summarize our data frame with the`summary()` function.  
```{r}
#summary()
```

`glimpse()` is another useful summary function.
```{r}
#glimpse()
```

`nrow()` gives the numbers of rows.
```{r}
#nrow()
```

`ncol` gives the number of columns.
```{r}
#ncol()
```

`names` gives the column names.
```{r}
#names()
```

`head()` prints the first n rows of the data frame.
```{r}
#head()
```

`tail()` prinst the last n rows of the data frame.
```{r}
#tail()
```

`table()` is useful when you have a limited number of categorical variables. It produces fast counts of the number of observations in a variable.  
```{r}
#table(data_frame$col_name)
```

We can also click on the data frame in the Environment tab or type View() to better see it.
```{r}
#View()
```

### Filter command

Filter is a way of pulling out observations that meet specific criteria in a variable. 
```{r}
#filter()
```

(Lab 4.2)
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a variable

`filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).


## Lab 4.1

### Select command
Select allows you to pull out columns of interest from a dataframe. To do this, just add the names of the columns to the `select()` command. The order in which you add them, will determine the order in which they appear in the output.
```{r}
#select()
```

There are also other ways you can reconfigure, modify, etc.:

To add a range of columns use `start_col:end_col`. This is super helpful when you have variables that are numbered

The - operator is useful in select. It allows us to select everything except the specified variables.
```{r}
#select(data_frame, -"unwanted_col")
```

Options to select columns based on a specific criteria include:  
1. ends_with() = Select columns that end with a character string  
2. contains() = Select columns that contain a character string  
3. matches() = Select columns that match a regular expression; 'regex'  
4. one_of() = Select columns names that are from a group of names

the following code is helpful when you know that a column contains a letter (such as "a") followed by a subsequent string (such as "er").  
```{r}
#select(data_frame, matches("a.+er")) 
```

You can also select columns based on the class of data, such as numeric data
```{r}
#select_if(data_frame, is.numeric)
```

To select all columns that are *not* a class of data, you need to add a `~`.
```{r}
#select_if(data_frame, ~!is.numeric(.)) # ! means not
```



Imported data frames often have a mix of lower and uppercase column names. Use `toupper()` or `tolower()` to fix this issue. I always try to use lowercase to keep things consistent.  
```{r}
#tolower(data_frame)
#or
#select_all(data_frame, tolower)
```

Other super helpful way to clean up column names is to load the 'janitor' package. This package could conflict with other ackages you might be using, however.
```{r}
#library(janitor) 
#clean_names(data.frame) #super handy trick for cleaning up column names
#janitor::clean_names()
```

When naming columns, blank spaces are often added. Here is a trick to remove these.  
```{r}
#select_all(data_frame, ~str_replace(., " ", "_"))
```


## lab 4.2

### Using `filter()` with multiple observations  
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
```{r}
#filter(data_frame, length %in% c(167, 175))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
#filter(fish, between(scalelength, 2.5, 2.55))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
#filter(fish, near(radii_length_mm, 2, tol = 0.2))
```

### Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. 

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met. same as '&' 
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met. 


```{r}
#?count
```


## Lab 5.1

'levels()' is used to show all the possible values within a given variable
```{r}
#levels(fish$lakeid) 
```


## Lab 5.2


### `arrange()`
The `arrange()` command is a bit like a sort command in excel. Note that the default is ascending order.  

To sort in decreasing order (large to small), wrap the variable name in `desc()`.

### `mutate()`  
Mutate allows us to create a new column from existing columns in a data frame. 

## `mutate_all()`
This last function is super helpful when cleaning data. With "wild" data, there are often mixed entries (upper and lowercase), blank spaces, odd characters, etc. These all need to be dealt with before analysis.  

Here is an example that changes all entries to lowercase (if present).  
```{r}
#data_frame %>%
  #mutate_all(tolower)
```

Using the across function we can specify individual columns.
```{r}
#mammals %>% 
  #mutate(across(c("order", "family"), tolower))
```

## `if_else()`
We will briefly introduce `if_else()` here because it allows us to use `mutate()` but not have the entire column affected in the same way. In a sense, this can function like find and replace in a spreadsheet program. With `ifelse()`, you first specify a logical statement, afterwards what needs to happen if the statement returns `TRUE`, and lastly what needs to happen if it's  `FALSE`.

`-999.00` is sometimes used as a placeholder for NA (but, is a really bad idea). We can use `if_else()` to replace `-999.00` with `NA`.  
```{r}
#mammals %>% 
  #select(genus, species, newborn) %>%
  #mutate(newborn_new = ifelse(newborn == -999.00, NA, newborn))%>% 
  #arrange(newborn)
```


##Lab 6.2

### `tabyl`
The `janitor` package has many awesome functions. Here is its version of `table` which not only produces counts but also percentages. Very handy! Let's use it to explore the proportion of good guys and bad guys in the `superhero_info` data.  
```{r}
#tabyl()
```

### Ratios
for ratios, can put '/' between two variables to get their ratio


## Lab 7.1

Alternative to 'summary()' is 'skim' (as a part of the skimr package)
```{r}
#skim()
```

'hist()'Histograms are also a quick way to check the output.
```{r}
#hist()
```

### `summarize()`
`summarize()` will produce summary statistics for a given variable in a data frame. Is often used to find the mean, mix, max, sd,total, etc. of variables all at once. Ex:
```{r}
```{r}
#msleep %>% 
  #filter(bodywt>200) %>% 
  #summarize(mean_sleep_lg=mean(sleep_total),
            #min_sleep_lg=min(sleep_total),
            #max_sleep_lg=max(sleep_total),
            #sd_sleep_lg=sd(sleep_total),
            #total=n())
```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. 

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

### `group_by()`
The `summarize()` function is most useful when used in conjunction with `group_by()`. ex: Although producing a summary of body weight for all of the mammals in the data set is helpful, what if we were interested in body weight by feeding ecology?
```{r}
#msleep %>%
  #group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable
  #summarize(min_bodywt = min(bodywt),
            #max_bodywt = max(bodywt),
            #mean_bodywt = mean(bodywt),
            #total=n())
```

```{r}
?n=n()
```

### Counts

`count()` is an easy way of determining how many observations you have within a column. It acts like a combination of `group_by()` and `n()`. ex:
```{r}
#penguins %>% 
  #count(island, sort = T) #sort=T sorts the column in descending order
```

You can also use `count()` across multiple variables.

For counts, I also like `tabyl()`

## Lab 7.2

## `across()`
What about using `filter()` and `select()` across multiple variables? There is a function in dplyr called `across()` which is designed to work across multiple variables.

By using `across()` we can reduce the clutter and make things cleaner instead of using 'summarise()'. ex: 
```{r}
#penguins %>%
  #summarize(across(c(species, island, sex), n_distinct))
```

This is very helpful for continuous variables.
```{r}
#penguins %>%
  #summarize(across(contains("mm"), mean, na.rm=T))
```

can also summarize across all variables. ex:
```{r}
#penguins %>%
  #summarise_all(mean, na.rm=T) #notice that there is no summary for categorical variables
```