---
title: "lab_4_warmup"
output: html_document
date: "2024-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Warm-up
1. In a new R Markdown document, load the tidyverse and a package called `palmerpenguins`.
2. What are the dimensions of the dataset `penguins`?
3. What are the names of the variables in the `penguins` dataset?
4. How many individuals were sampled on each island in the `penguins` dataset?
5. What is the mean body mass for all individuals in the `penguins` dataset?


## 1. Load the libraries
```{r}
library("tidyverse")
library(palmerpenguins)
```

dont need to use read.csv because we already imported the data

```{r}
penguins
```


## 2. What are the dimensions of 'penguins'?
```{r}
dim(penguins)
```

## 3. What are the names of the variables (columns)?
```{r}
names(penguins)
```
 
 or can also do
```{r}
glimpse(penguins)
```
## 4. How many individuals are sampled on each island?
```{r}
penguins <- penguins
```
 (this stores 'penguins' as an object)
 
```{r}
table(penguins$island)
```
 
 or can also do
```{r}
summary(penguins$island)
```
 
```{r}
mean(penguins$body_mass_g, na.rm=T)
```
(include 'na.rm=T' because there are 'NA's' in the data)
